Think deep for reef resilience

### 2025-06-05 paper/paper_draft.qmd
Edit citation formatting and add config file for paper-specific sizes

Climate change is impacting species survival, persistence and ecosystem functions globally, including coral. Coral reefs are vulnerable to increasing magnitude and variation in environmental stressors such as ocean temperature, acidification and nutrient loading ([@foster2021]; [@gattuso2014]). Mass bleaching and mortality events linked to climate change can cause reef-building coral population declines, possibly leading to long-term and large-scale coral-macroalgae regime shifts ([@arif2022]; [@wild2011]). Climate impacts on coral reefs are, and will continue to, decrease the ability of coral reefs to provide ecological functions such as benthic habitat and nutrient cycling ([@forster2021]; [@williams2019]) as well as ecosystem services such as food, protection, tourism and Indigenous cultural value ([@eddy2021]; [@stoeckl2021]). We aim to assess the impacts of possible climate futures and identify the characteristics of resilient reefs using the latest generation of larval connectivity research with scenario-based population modelling for the Great Barrier Reef ecosystem.

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of thousands of individual reefs. They are interconnected through coral and fish larval dispersal patterns mediated by hydrodynamic, spatial, and ecological processes ([@grimaldi2022]). Severe declines in coral cover are projected due to compounding impacts such as ocean warming and acidification, intensification of Tropical Cyclones, and Crown of Thorns Starfish outbreaks ([@death2012]; [@fox-kemper2021]; [@seneviratne2021]; [@sweatman2011]; [@wolff2018]). Ominously, the 2023-24 Summer mass bleaching event ([@cantin2024]) may be signalling a shift towards this reef future, with some level of bleaching recorded on 79% of surveyed reefs.

Environmental abiotic factors influence how reefs will experience climate change pressures, including the spatio-temporal patterns of thermal stress itself, and indirect factors that influence the stress experienced by corals. Environmental factors such as tropical cyclones and wind driven mixing ([@lin2023]; [@mcclanahan2005]); currents ([@grimaldi2022]) and depth ([@álvarez-noriega2018]; [@baird2018]; [@smith2014]) can decrease the magnitude of thermal stress corals experience. On the other hand, high light intensity and nutrient levels in the benthic zone can increase coral bleaching ([@decarlo2020]; [@higuchi2015]). These environmental factors influence the resilience of reefs to climate change pressures in combination with related biological factors such as larval connectivity and thermal tolerance.

Connectivity in systems such as the GBR is an important contributor to system resilience, as strong connections are required to repopulate locations after disturbance, and drive adaptation to climate pressures ([@angeler2023]; [@elmhirst2009]; [@hock2017]). When populations that are strong larval suppliers to external reefs are impacted by disturbances, increased adult coral mortality will cause decreased outgoing larvae in the following years. Weaker larval connections then slow the recovery of the externally connected locations ([@figueiredo2022]; [@hock2017]; [@munday2009]). Populations with greater strength and number of outgoing than incoming larval connections - herein referred to as larval sources - have lower resilience to pressures than populations with greater incoming than outgoing larval flows - referred to as larval sinks - as larval source populations lack external sources of larvae that are important in contributing to reef recovery and recruitment of juveniles after disturbances to existing populations ([@ani2024]; [@hock2017]). Larval source populations are an important component of overall reef resilience as they can contribute larvae to recovering populations at other locations ([@hock2017]; [@thomas2015]). Connectivity will be altered by future climate change impacts such as population fragmentation, increasing larval mortality, and changes to coastal hydrodynamics ([@figueiredo2022]; [@munday2009]), and influences the biology of surrounding reef populations.

Many additional biological factors including coral size ([@brandt2009]), taxa ([@álvarez-noriega2018]; [@mcclanahan2005]), and acclimation and inherited thermal tolerance ([@humanes2022]; [@lachs2024]), influence the outcome of high coral heat stress. It has been shown that corals have the adaptive capacity to develop thermal tolerance levels capable of delaying the onset of widespread bleaching under low emissions scenarios ([@bairos-novak2021]; [@humanes2022]).

Incorporating the latest advancements in whole ecosystem larval connectivity modelling and coral thermal response and adaptation research into coral population modelling allows us to investigate how these abiotic and biotic factors interact and will influence coral climate change resilience into the future. Improvements in coral reef climate response modelling, including sampling from a range of possible climate futures, allow for fine-scale timeseries analyses of uncertain climate change and ecological scenarios ([@bode2017]; [@chollett2022]; [@donner2018]; [@klein2024]; [@uusitalo2015]).

@@

## Environmental modelling context
![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all considered reefs (points). The green polygon represents the Far Northern management area, the yellow polygon represents the Cairns/Cooktown management area, the pink polygon represents the Townsville/Whitsunday management area and the blue polygon represents the Mackay/Capricorn management area. The centroid of each reef is represented as a point for visual clarity, and is coloured according to its GBRMPA reefal bioregion. An annotation represents the approximate location of the Australian Institute of Marine Science Townsville site at Cape Cleveland.](/../outputs/ADRIA_results/HighResCoralStress/figs/region_map.png){#fig-region-map width=`{julia} map_width`}

The spatial extent of GBR coral reefs used in this investigation includes 3806 reef areas. It follows the reef definitions as adopted in ReefMod-GBR ([@bozec2022]) (@fig-region-map). Annually and spatially averaged Degree Heating Week (DHW) data is used to represent future thermal stress at each reef location. Five Global Climate Models (GCMs) (EC-Earth3-Veg, ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GDFL-CM4) from the Sixth Phase of the Coupled Model Intercomparison Project (CMIP6) were selected for this study. Statistically downscaled 1km resolution Sea Surface Temperature (SST) data for the Great Barrier Reef were obtained from the HighResCoralStress database ([@dixon2022][@dixon]). Maximum monthly mean DHW levels were calculated from the SST data following NOAA methodology ([@liu2017]; [@skirving2020]). These GCMs were selected from available GCMs downscaled by Dixon et al. [-@dixon] based on evaluation by Grose et al. [-@grose2023], which found high correlation between observed temperature, rainfall, El Niño Southern Oscillation and Indian Ocean Dipole patterns in the Australian terrestrial region, and the corresponding mechanisms in these GCMs. Equilibrium climate sensitivity (ECS) is defined, by the IPCC, as the long term warming caused by a doubling of carbon dioxide above its pre-industrial concentration ([@forster2021]). The four GCMs span a range of ECS values within the IPCC assessed likely range, and GCM development families, allowing for a range of possible climate futures to be explored, capturing the uncertainty in future responses to CO2 increases (@fig-ecs; [@grose2023]; [@shepherd2018]). The following analyses are performed using each GCM separately. SSP 2-4.5 is used as a plausible emission pathway under current emission trajectories, with similar atmospheric CO2 levels remaining until 2050 ([@hausfather2020]; IPCC [-@climate2023]).

@@

CoralBlox is a coral population growth and mortality model that simulates groups of corals growing in diameter and competing for limited space over time ([@ribeirodealmeida2024]). This model incorporates linear extension (growth) and background mortality rates that are dependent on the size and functional group of a coral colony. Environmental disturbances and reproduction are not directly considered within CoralBlox. Instead, these factors are considered by making changes to the distribution of colonies within CoralBlox timesteps using the ADRIA (Adaptive Dynamic Reef Intervention Algorithms) platform ([@iwanaga2023]). Within ADRIA, we incorporate coral life history and environmental events in the following order: coral colony growth; coral reproduction; larval connectivity and settlement; and coral bleaching due to heat stress. As ADRIA and CoralBlox are coupled together to represent the system, we refer to both as ADRIA-CoralBlox.

@@

Timeseries analysis is performed by first grouping reefs into three different scales, GBRMPA reefal bioregion, GBRMPA management area and GBR-wide, to investigate the consistency of cluster identification. To identify groups of reefs at each analysis scale that are resilient, intermediate and non-resilient we clustered reef relative coral cover timeseries into three clusters at each scale. To quantify the behaviour of timeseries and identify timeseries that behave similarly we calculate the complexity-invariant distance (CID) between each timeseries. The complexity-invariant distance refers to the Euclidean distance between two timeseries multiplied (corrected) by the ratio of the complexities of the timeseries ([@batista2014]). These values create a distance matrix of n * n dimensions (where n is the number of timeseries in the group) that is clustered using a k-medoids clustering algorithm ([@steinmann2020]).

@@

At management area analysis scales, there are not consistent substantial differences in source – to – sink ratios (@sfig-access-esm1-5-manarea-logsosi; @sfig-access-cm2-manarea-logsosi; @sfig-ec-earth3-veg-manarea-logsosi; @sfig-gfdl-cm4-manarea-logsosi; @sfig-noresm2-mm-manarea-logsosi) and total connectivity strength (@sfig-access-esm1-5-manarea-totalconnectivity; @sfig-access-cm4-manarea-totalconnectivity; @sfig-ec-earth3-veg-manarea-totalconnectivity; @sfig-gfdl-cm4-manarea-totalconnectivity; @sfig-noresm2-mm-manarea-totalconnectivity).

@@

The clusters identified with timeseries clustering at the GBR scale show similar patterns to clusters identified at bioregion and management area scales. Reef depth (@fig-access-esm1-5-gbr-depth; @sfig-access-cm2-gbr-depth; @sfig-ec-earth3-veg-gbr-depth; @sfig-gfdl-cm4-gbr-depth; @sfig-noresm2-mm-gbr-depth) and total cover – DHW correlation (@sfig-access-esm1-5-gbr-dhwcor; @sfig-access-cm2-gbr-dhwcor; @sfig-ec-earth3-veg-gbr-dhwcor; @sfig-gfdl-cm4-gbr-dhwcor; @sfig-noresm2-mm-gbr-dhwcor) decrease, while the proportion of coral-habitable area initially occupied by corals increases (@sfig-access-esm1-5-gbr-initialproportion; @sfig-access-cm2-gbr-initialproportion; @sfig-ec-earth3-veg-gbr-initialproportion; @sfig-gfdl-cm4-gbr-initialproportion; @sfig-noresm2-mm-gbr-initialproportion), as the clusters decrease from ‘high’ to ‘low’ cover levels. There are no substantial differences between the clusters in mean DHW (@sfig-access-esm1-5-gbr-dhwmean; @sfig-access-cm4-gbr-dhwmean; @sfig-ec-earth3-veg-gbr-dhwmean; @sfig-gfdl-cm4-gbr-dhwmean; @sfig-noresm2-mm-gbr-dhwmean) and DHW tolerance (@sfig-access-esm1-5-gbr-dhwtol; @sfig-access-cm2-gbr-dhwtol; @sfig-ec-earth3-veg-gbr-dhwtol; @sfig-gfdl-cm4-gbr-dhwtol; @sfig-noresm2-mm-gbr-dhwtol) levels of reefs. Similarly, there are no substantial differences between clusters in connectivity metrics source – to – sink ratio and total connectivity strength (@sfig-access-esm1-5-gbr-logsosi; @sfig-access-cm2-gbr-logsosi; @sfig-ec-earth3-veg-gbr-logsosi; @sfig-gfdl-cm4-gbr-logsosi; @sfig-noresm2-mm-gbr-logsosi; @sfig-access-esm1-5-gbr-totalconnectivity; @sfig-access-cm2-gbr-totalconnectivity; @sfig-ec-earth3-veg-gbr-totalconnectivity; @sfig-gfdl-cm4-gbr-totalconnectivity; @sfig-noresm2-mm-gbr-totalconnectivity).

@@

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced, and exhibit overall similar DHW timeseries. While these reefs experience similar magnitude DHWs based on SST, the proportion of the reef’s coral population impacted by heat stress events depends on the depth of the reef. Analysis from Baird et al. [-@baird2018] is used to calculate the degree of protection provided to a reef based on the median depth. Our results suggest that within shallow reef ecosystems (depth < 30m), depth contributes greatly to reef resilience and the ability of the reef to recover following heat stress. While similar findings have been demonstrated in the eastern Pacific ([@smith2014]), depth refugia analyses have often focussed on the potential of mesophotic reefs (> 30m deep) as thermal refugia ([@bongaerts2010]; [@holstein2015]; [@semmler2017]). In contrast, Venegas et al. [-@venegas2019] found no association between depth and the alleviation of heat stress across a range of shallow reefs in the Pacific. The depth refugia found on a reef in turn influences the degree of correlation between coral cover timeseries and DHW levels. Reefs in low and medium clusters have lower levels of cover – DHW correlation than those in high clusters, due to their initial sharp declines in coral cover combined with low depth – refugia in the face of persistently high thermal stress these reefs do not recover post 2030, leaving little ability to decline further in periods of high thermal stress and recover in periods of low thermal stress.

@@

Current reef restoration and intervention-investigation efforts on the Great Barrier Reef, such as deployment of coral seeding and fragment devices target reefs with suitable restoration sites in the 1-10m depth range ([@howlett2021]; [@randall2021]; [@whitman2024]). One of the main depth constraints is being able to visually identify the target site from the surface for deployment and ongoing monitoring, in combination with factors such as deployment method and tourism return value. Our results suggest that of the reefs analysed within the 1-10m depth range (1800) 87% (1569) of reefs fall into the low and medium clusters (at the Great Barrier Reef scale), with many suffering large future declines in reef health under climate change and possibly limiting the potential of interventions to deliver recovery in these locations. Reef intervention planning is a complex and challenging decision making process ([@anthony2020]), and our results suggest that deployment site depth is likely an important factor to consider when assessing the long-term resilience of reef sites.

Overall, our findings do not suggest that resilient reefs in ‘high’ cover clusters have greater thermal tolerance than reefs in lower coral cover clusters. This is in contrast with literature suggesting corals have the capacity to adapt to increasing climate change pressures ([@humanes2022]; [@lachs2024]), including the investigation by Bairos-Novak et al. [-@bairos-novak2021] that revealed high heritability of bleaching and survival response traits used to inform the distributions of thermal tolerance traits used in this investigation. The lack of influence of thermal tolerance on reef outcomes is likely due to reef being impacted by overall similar levels of thermal stress, combined with the greater importance of factors such as reef depth and initial coral cover.

It has been suggested that a primary facet of reef resilience is the degree of larval connectivity between reefs and the supply of recruits that a reef receives following disturbance ([@hock2017]). It was expected that resilient reefs would have greater incoming than outgoing larval connectivity to promote recovery in years following thermal stress, while non-resilient reefs would have greater outgoing than incoming connectivity inhibiting recruitment of corals ability post thermal stress. In contrast, our results suggest that there are few differences between resilient and non-resilient reefs in connectivity metrics such as the ratio of outgoing to incoming larval connectivity and total connectivity strength. Hock et al. ([-@hock2017]) used larval connectivity modelling to identify resilient reefs that may supply up to 47% of the wider Great Barrier Reef ecosystem with larvae following disturbance. A critique of this approach ([@bode2018]) is that results obtained are highly sensitive to the biophysical dispersal model used, limiting the application of specific identified reefs. Reef depth was not considered in these investigations of reef network resilience. These investigations were performed using prior generations of larval connectivity models parameterised for 2008-2013 ([@hock2017]) and 1996-2002 ([@bode2018]) time periods respectively. Using updated 1km resolution larval connectivity across the whole GBR, our results suggest that larval connectivity alone is not as important for reef resilience and recovery as environmental refugia from thermal stress such as depth.

Five different GCMs were included in analyses to capture uncertainty in climate system responses to increasing atmospheric carbon dioxide levels. The large differences in relative cover timeseries seen across the GCMs suggest that it is important to consider this uncertainty in analyses of biological climate change responses. This is supported by recent literature suggesting an effective approach to capturing climate change uncertainty is to model climate change responses using a range of climate models separately ([@shepherd2018]). However, the patterns identified between low, medium and high clusters of reefs suggest that the importance of depth refugia is consistent across scenarios. Similarly, at the management area scale, only 1% of reefs were assigned a ‘low’ and ‘high’ cluster in different GCMs, suggesting that intrinsic reef properties can prevent high cover resilient reefs from crashing under a different future climate scenario.

There are multiple caveats in how coral ecosystem dynamics are modelled in this investigation. We model coral colonies as objects with circular diameter values and track these diameters over timesteps through growth. Due to the availability of quality future projected data we do not consider the impacts of waves, cyclones and currents on coral colony survival and bleaching (insert citations for waves affecting survival and bleaching). We do not consider the impacts of Crown of Thorns Starfish (CoTS) on coral survival and reef resilience (insert citation for cots affecting corals). Furthermore, there are multiple caveats to acknowledge in the data sources used in this investigation. Uncertainty in connectivity modelling outputs, satellite-derived bathymetry data and future thermal stress projections are likely to influence the results obtained in this investigation. To limit the influence of bathymetry data on results, reefs with less than 5% data coverage are removed from analyses. We investigate a range of future climate responses and thermal projections, and our results suggest the factors that influence reef resilience are consistent across the tested futures.

Further investigations could...

@@

Overall, our results suggest that when modelled as part of a large and complex ecosystem, reef depth is a key factor contributing to ongoing reef resilience. We provide evidence in support of depth as a form of refugia for shallow reefs. While protecting ‘key node’ reefs in larval connectivity networks is an important consideration for reef restoration efforts ([@anthony2020]; [@hock2017]), our results suggest that resilient reefs do not have greater incoming or total connectivity to contribute to their resilience. Instead, we suggest that depth, and likely other environmental thermal refugia mechanisms contribute greater to overall reef resilience and recovery. Further investigation could utilise the scenario modelling and dynamic intervention capabilities of the ADRIA platform to explore the effectiveness of prioritising reef depth compared to larval connectivity in reef intervention decision making.

@@

## Model parameterisation
CoralBlox uses user-defined size classes that influence size-dependent growth and mortality rates, through which the coral colonies progress as they grow in diameter. Within ADRIA, we define the distributions of these parameters, and the size classes based on ecological monitoring data from the Ecological Intelligence for Reef Restoration Program (EcoRRAP) ([@australian-institute-of-marine-science2025]; [@ribeirodealmeida2024]).

>>>>

Climate change is impacting species survival, persistence and ecosystem functions globally, including coral. Coral reefs are vulnerable to increasing magnitude and variation in environmental stressors such as ocean temperature, acidification and nutrient loading [@foster2021; @gattuso2014]. Mass bleaching and mortality events linked to climate change can cause reef-building coral population declines, possibly leading to long-term and large-scale coral-macroalgae regime shifts [@arif2022; @wild2011]. Climate impacts on coral reefs are, and will continue to, decrease the ability of coral reefs to provide ecological functions such as benthic habitat and nutrient cycling [@forster2021; @williams2019] as well as ecosystem services such as food, protection, tourism and Indigenous cultural value [@eddy2021; @stoeckl2021]. We aim to assess the impacts of possible climate futures and identify the characteristics of resilient reefs using the latest generation of larval connectivity research with scenario-based population modelling for the Great Barrier Reef ecosystem.

The Great Barrier Reef (GBR) is the world's largest coral reef ecosystem consisting of thousands of individual reefs. They are interconnected through coral and fish larval dispersal patterns mediated by hydrodynamic, spatial, and ecological processes [@grimaldi2022]. Severe declines in coral cover are projected due to compounding impacts such as ocean warming and acidification, intensification of Tropical Cyclones, and Crown of Thorns Starfish outbreaks [@death2012; @fox-kemper2021; @seneviratne2021; @sweatman2011; @wolff2018]. Ominously, the 2023-24 Summer mass bleaching event ([@cantin2024]) may be signalling a shift towards this reef future, with some level of bleaching recorded on 79% of surveyed reefs.

Environmental abiotic factors influence how reefs will experience climate change pressures, including the spatio-temporal patterns of thermal stress itself, and indirect factors that influence the stress experienced by corals. Environmental factors such as tropical cyclones and wind driven mixing [@lin2023; @mcclanahan2005]; currents [@grimaldi2022] and depth [@álvarez-noriega2018; @baird2018; @smith2014] can decrease the magnitude of thermal stress corals experience. On the other hand, high light intensity and nutrient levels in the benthic zone can increase coral bleaching [@decarlo2020; @higuchi2015]. These environmental factors influence the resilience of reefs to climate change pressures in combination with related biological factors such as larval connectivity and thermal tolerance.

Connectivity in systems such as the GBR is an important contributor to system resilience, as strong connections are required to repopulate locations after disturbance, and drive adaptation to climate pressures [@angeler2023; @elmhirst2009; @hock2017]. When populations that are strong larval suppliers to external reefs are impacted by disturbances, increased adult coral mortality will cause decreased outgoing larvae in the following years. Weaker larval connections then slow the recovery of the externally connected locations [@figueiredo2022; @hock2017; @munday2009]. Populations with greater strength and number of outgoing than incoming larval connections - herein referred to as larval sources - have lower resilience to pressures than populations with greater incoming than outgoing larval flows - referred to as larval sinks - as larval source populations lack external sources of larvae that are important in contributing to reef recovery and recruitment of juveniles after disturbances to existing populations [@ani2024; @hock2017]. Larval source populations are an important component of overall reef resilience as they can contribute larvae to recovering populations at other locations [@hock2017; @thomas2015]. Connectivity will be altered by future climate change impacts such as population fragmentation, increasing larval mortality, and changes to coastal hydrodynamics [@figueiredo2022; @munday2009], and influences the biology of surrounding reef populations.

Many additional biological factors including coral size [@brandt2009], taxa [@álvarez-noriega2018; @mcclanahan2005], and acclimation and inherited thermal tolerance[@humanes2022; @lachs2024], influence the outcome of high coral heat stress. It has been shown that corals have the adaptive capacity to develop thermal tolerance levels capable of delaying the onset of widespread bleaching under low emissions scenarios [@bairos-novak2021; @humanes2022].

Incorporating the latest advancements in whole ecosystem larval connectivity modelling and coral thermal response and adaptation research into coral population modelling allows us to investigate how these abiotic and biotic factors interact and will influence coral climate change resilience into the future. Improvements in coral reef climate response modelling, including sampling from a range of possible climate futures, allow for fine-scale timeseries analyses of uncertain climate change and ecological scenarios [@bode2017; @chollett2022; @donner2018; @klein2024; @uusitalo2015].

@@

## Environmental modelling context

![Map of the Great Barrier Reef region, including GBRMPA management areas (polygons) and all considered reefs (points). The green polygon represents the Far Northern management area, the yellow polygon represents the Cairns/Cooktown management area, the pink polygon represents the Townsville/Whitsunday management area and the blue polygon represents the Mackay/Capricorn management area. The centroid of each reef is represented as a point for visual clarity, and is coloured according to its GBRMPA reefal bioregion. An annotation represents the approximate location of the Australian Institute of Marine Science Townsville site at Cape Cleveland.](/../outputs/ADRIA_results/HighResCoralStress/figs/region_map.png){#fig-region-map width=`{julia} map_width`}

The spatial extent of GBR coral reefs used in this investigation includes 3806 reef areas. It follows the reef definitions as adopted in ReefMod-GBR [@bozec2022] (@fig-region-map). Annually and spatially averaged Degree Heating Week (DHW) data is used to represent future thermal stress at each reef location. Five Global Climate Models (GCMs) (EC-Earth3-Veg, ACCESS-CM2, ACCESS-ESM1-5, NorESM2-MM and GDFL-CM4) from the Sixth Phase of the Coupled Model Intercomparison Project (CMIP6) were selected for this study. Statistically downscaled 1km resolution Sea Surface Temperature (SST) data for the Great Barrier Reef were obtained from the HighResCoralStress database [@dixon2022; @dixon]. Maximum monthly mean DHW levels were calculated from the SST data following NOAA methodology [@liu2017; @skirving2020]. These GCMs were selected from available GCMs downscaled by Dixon et al. [-@dixon] based on evaluation by Grose et al. [-@grose2023], which found high correlation between observed temperature, rainfall, El Niño Southern Oscillation and Indian Ocean Dipole patterns in the Australian terrestrial region, and the corresponding mechanisms in these GCMs. Equilibrium climate sensitivity (ECS) is defined, by the IPCC, as the long term warming caused by a doubling of carbon dioxide above its pre-industrial concentration [@forster2021]. The four GCMs span a range of ECS values within the IPCC assessed likely range, and GCM development families, allowing for a range of possible climate futures to be explored, capturing the uncertainty in future responses to CO2 increases (@fig-ecs; [@grose2023; @shepherd2018]). The following analyses are performed using each GCM separately. SSP 2-4.5 is used as a plausible emission pathway under current emission trajectories, with similar atmospheric CO2 levels remaining until 2050 [@hausfather2020; IPCC -@climate2023].

@@

CoralBlox is a coral population growth and mortality model that simulates groups of corals growing in diameter and competing for limited space over time [@ribeirodealmeida2024]. This model incorporates linear extension (growth) and background mortality rates that are dependent on the size and functional group of a coral colony. Environmental disturbances and reproduction are not directly considered within CoralBlox. Instead, these factors are considered by making changes to the distribution of colonies within CoralBlox timesteps using the ADRIA (Adaptive Dynamic Reef Intervention Algorithms) platform [@iwanaga2023]. Within ADRIA, we incorporate coral life history and environmental events in the following order: coral colony growth; coral reproduction; larval connectivity and settlement; and coral bleaching due to heat stress. As ADRIA and CoralBlox are coupled together to represent the system, we refer to both as ADRIA-CoralBlox.

@@

Timeseries analysis is performed by first grouping reefs into three different scales, GBRMPA reefal bioregion, GBRMPA management area and GBR-wide, to investigate the consistency of cluster identification. To identify groups of reefs at each analysis scale that are resilient, intermediate and non-resilient we clustered reef relative coral cover timeseries into three clusters at each scale. To quantify the behaviour of timeseries and identify timeseries that behave similarly we calculate the complexity-invariant distance (CID) between each timeseries. The complexity-invariant distance refers to the Euclidean distance between two timeseries multiplied (corrected) by the ratio of the complexities of the timeseries [@batista2014]. These values create a distance matrix of n * n dimensions (where n is the number of timeseries in the group) that is clustered using a k-medoids clustering algorithm [@steinmann2020].

@@

At management area analysis scales, there are not consistent substantial differences in source – to – sink ratios (@sfig-access-esm1-5-manarea-logsosi; @sfig-access-cm2-manarea-logsosi; @sfig-ec-earth3-veg-manarea-logsosi; @sfig-gfdl-cm4-manarea-logsosi; @sfig-noresm2-mm-manarea-logsosi) and total connectivity strength (@sfig-access-esm1-5-manarea-totalconnectivity; @sfig-access-cm2-manarea-totalconnectivity; @sfig-ec-earth3-veg-manarea-totalconnectivity; @sfig-gfdl-cm4-manarea-totalconnectivity; @sfig-noresm2-mm-manarea-totalconnectivity).

@@

The clusters identified with timeseries clustering at the GBR scale show similar patterns to clusters identified at bioregion and management area scales. Reef depth (@fig-access-esm1-5-gbr-depth; @sfig-access-cm2-gbr-depth; @sfig-ec-earth3-veg-gbr-depth; @sfig-gfdl-cm4-gbr-depth; @sfig-noresm2-mm-gbr-depth) and total cover – DHW correlation (@sfig-access-esm1-5-gbr-dhwcor; @sfig-access-cm2-gbr-dhwcor; @sfig-ec-earth3-veg-gbr-dhwcor; @sfig-gfdl-cm4-gbr-dhwcor; @sfig-noresm2-mm-gbr-dhwcor) decrease, while the proportion of coral-habitable area initially occupied by corals increases (@sfig-access-esm1-5-gbr-initialproportion; @sfig-access-cm2-gbr-initialproportion; @sfig-ec-earth3-veg-gbr-initialproportion; @sfig-gfdl-cm4-gbr-initialproportion; @sfig-noresm2-mm-gbr-initialproportion), as the clusters decrease from ‘high’ to ‘low’ cover levels. There are no substantial differences between the clusters in mean DHW (@sfig-access-esm1-5-gbr-dhwmean; @sfig-access-cm2-gbr-dhwmean; @sfig-ec-earth3-veg-gbr-dhwmean; @sfig-gfdl-cm4-gbr-dhwmean; @sfig-noresm2-mm-gbr-dhwmean) and DHW tolerance (@sfig-access-esm1-5-gbr-dhwtol; @sfig-access-cm2-gbr-dhwtol; @sfig-ec-earth3-veg-gbr-dhwtol; @sfig-gfdl-cm4-gbr-dhwtol; @sfig-noresm2-mm-gbr-dhwtol) levels of reefs. Similarly, there are no substantial differences between clusters in connectivity metrics source – to – sink ratio and total connectivity strength (@sfig-access-esm1-5-gbr-logsosi; @sfig-access-cm2-gbr-logsosi; @sfig-ec-earth3-veg-gbr-logsosi; @sfig-gfdl-cm4-gbr-logsosi; @sfig-noresm2-mm-gbr-logsosi; @sfig-access-esm1-5-gbr-totalconnectivity; @sfig-access-cm2-gbr-totalconnectivity; @sfig-ec-earth3-veg-gbr-totalconnectivity; @sfig-gfdl-cm4-gbr-totalconnectivity; @sfig-noresm2-mm-gbr-totalconnectivity).

@@

Reefs in different clusters largely do not differ in mean modelled DHW levels experienced, and exhibit overall similar DHW timeseries. While these reefs experience similar magnitude DHWs based on SST, the proportion of the reef’s coral population impacted by heat stress events depends on the depth of the reef. Analysis from Baird et al. [-@baird2018] is used to calculate the degree of protection provided to a reef based on the median depth. Our results suggest that within shallow reef ecosystems (depth < 30m), depth contributes greatly to reef resilience and the ability of the reef to recover following heat stress. While similar findings have been demonstrated in the eastern Pacific [@smith2014], depth refugia analyses have often focussed on the potential of mesophotic reefs (> 30m deep) as thermal refugia [@bongaerts2010; @holstein2015; @semmler2017]. In contrast, Venegas et al. [-@venegas2019] found no association between depth and the alleviation of heat stress across a range of shallow reefs in the Pacific. The depth refugia found on a reef in turn influences the degree of correlation between coral cover timeseries and DHW levels. Reefs in low and medium clusters have lower levels of cover – DHW correlation than those in high clusters, due to their initial sharp declines in coral cover combined with low depth – refugia in the face of persistently high thermal stress these reefs do not recover post 2030, leaving little ability to decline further in periods of high thermal stress and recover in periods of low thermal stress.

@@

Current reef restoration and intervention-investigation efforts on the Great Barrier Reef, such as deployment of coral seeding and fragment devices target reefs with suitable restoration sites in the 1-10m depth range [@howlett2021; @randall2021; @whitman2024]. One of the main depth constraints is being able to visually identify the target site from the surface for deployment and ongoing monitoring, in combination with factors such as deployment method and tourism return value. Our results suggest that of the reefs analysed within the 1-10m depth range (1800) 87% (1569) of reefs fall into the low and medium clusters (at the Great Barrier Reef scale), with many suffering large future declines in reef health under climate change and possibly limiting the potential of interventions to deliver recovery in these locations. Reef intervention planning is a complex and challenging decision making process [@anthony2020], and our results suggest that deployment site depth is likely an important factor to consider when assessing the long-term resilience of reef sites.

Overall, our findings do not suggest that resilient reefs in ‘high’ cover clusters have greater thermal tolerance than reefs in lower coral cover clusters. This is in contrast with literature suggesting corals have the capacity to adapt to increasing climate change pressures [@humanes2022; @lachs2024], including the investigation by Bairos-Novak et al. [-@bairos-novak2021] that revealed high heritability of bleaching and survival response traits used to inform the distributions of thermal tolerance traits used in this investigation. The lack of influence of thermal tolerance on reef outcomes is likely due to reef being impacted by overall similar levels of thermal stress, combined with the greater importance of factors such as reef depth and initial coral cover.

It has been suggested that a primary facet of reef resilience is the degree of larval connectivity between reefs and the supply of recruits that a reef receives following disturbance [@hock2017]. It was expected that resilient reefs would have greater incoming than outgoing larval connectivity to promote recovery in years following thermal stress, while non-resilient reefs would have greater outgoing than incoming connectivity inhibiting recruitment of corals ability post thermal stress. In contrast, our results suggest that there are few differences between resilient and non-resilient reefs in connectivity metrics such as the ratio of outgoing to incoming larval connectivity and total connectivity strength. Hock et al. [-@hock2017] used larval connectivity modelling to identify resilient reefs that may supply up to 47% of the wider Great Barrier Reef ecosystem with larvae following disturbance. A critique of this approach [@bode2018] is that results obtained are highly sensitive to the biophysical dispersal model used, limiting the application of specific identified reefs. Reef depth was not considered in these investigations of reef network resilience. These investigations were performed using prior generations of larval connectivity models parameterised for 2008-2013 [@hock2017] and 1996-2002 [@bode2018] time periods respectively. Using updated 1km resolution larval connectivity across the whole GBR, our results suggest that larval connectivity alone is not as important for reef resilience and recovery as environmental refugia from thermal stress such as depth.

Five different GCMs were included in analyses to capture uncertainty in climate system responses to increasing atmospheric carbon dioxide levels. The large differences in relative cover timeseries seen across the GCMs suggest that it is important to consider this uncertainty in analyses of biological climate change responses. This is supported by recent literature suggesting an effective approach to capturing climate change uncertainty is to model climate change responses using a range of climate models separately [@shepherd2018]. However, the patterns identified between low, medium and high clusters of reefs suggest that the importance of depth refugia is consistent across scenarios. Similarly, at the management area scale, only 1% of reefs were assigned a ‘low’ and ‘high’ cluster in different GCMs, suggesting that intrinsic reef properties can prevent high cover resilient reefs from crashing under a different future climate scenario.

There are multiple caveats in how coral ecosystem dynamics are modelled in this investigation. We model coral colonies as objects with circular diameter values and track these diameters over timesteps through growth. Due to the availability of quality future projected data we do not consider the impacts of waves, cyclones and currents on coral colony survival and bleaching (insert citations for waves affecting survival and bleaching). We do not consider the impacts of Crown of Thorns Starfish (CoTS) on coral survival and reef resilience (insert citation for cots affecting corals). Furthermore, there are multiple caveats to acknowledge in the data sources used in this investigation. Uncertainty in connectivity modelling outputs, satellite-derived bathymetry data and future thermal stress projections are likely to influence the results obtained in this investigation. To limit the influence of bathymetry data on results, reefs with less than 5% data coverage are removed from analyses. We investigate a range of future climate responses and thermal projections to account for a range of possible futures, and our results suggest the factors that influence reef resilience are consistent across these scenarios.

Further analyses could build on the findings of this study by analysing the impacts of prioritising reef depth in reef restoration intervention decision making. The ADRIA platform is developed as a modelling tool to incorporate multi-criteria decision analysis (MCDA) and ecosystem scenario modelling into reef restoration decision analysis. The platform allows for the investigation of different reef characteristics to be targetted in decision making and models the evolving outcomes as a result of reef interventions at various locations. The capability of ADRIA-CoralBlox could be further expanded by incorporating the impacts of waves, cyclones, CoTS and currents into reef resilience investigations [@lin2023; @mcclanahan2005; @grimaldi2022; @liu2024].

@@

Overall, our results suggest that when modelled as part of a large and complex ecosystem, reef depth is a key factor contributing to ongoing reef resilience. We provide evidence in support of depth as a form of refugia for shallow reefs. While protecting ‘key node’ reefs in larval connectivity networks is an important consideration for reef restoration efforts [@anthony2020; @hock2017], our results suggest that resilient reefs do not have greater incoming or total connectivity to contribute to their resilience. Instead, we suggest that depth, and likely other environmental thermal refugia mechanisms contribute greater to overall reef resilience and recovery. We suggest further analyses that investigate the benefits of incorporating depth-based resilience into reef decision making.

@@

## Model parameterisation

CoralBlox uses user-defined size classes that influence size-dependent growth and mortality rates, through which the coral colonies progress as they grow in diameter. Within ADRIA, we define the distributions of these parameters, and the size classes based on ecological monitoring data from the Ecological Intelligence for Reef Restoration Program (EcoRRAP) ([@australianinstituteofmarinescience2025; @ribeirodealmeida2024]).

## ACCESS-ESM1-5

### Bioregion scale
